---
title: "Do Income and Population Affect Crimes Committed in Neighbourhoods? An Analysis of the 2020 Toronto Crime Data"
#subtitle: "NA"
author: 
  - Janel Gilani
thanks: "Code and data are available at: https://github.com/JanelGilani/spread-of-crimes-in-toronto"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| message: false
#| echo: false
#| include: false
#| warning: false

#### Workspace set-up ####

## Load packages
library(knitr)
library(kableExtra)
library(here)
library(ggplot2)
library(ggbeeswarm)
library(ggrepel)
library(tidyverse)

#### Read in cleaned data ####

cleaned_crimes_data <- read_csv(file = here("outputs/data/crimes_cleaned_data.csv"), show_col_types = FALSE)
cleaned_neighbourhood_data <- read_csv(file = here("outputs/data/neighbourhood_cleaned_data.csv"), show_col_types = FALSE)
final_neighbourhood_data <- read_csv(file = here("outputs/data/hood_crime_analysis_data.csv"), show_col_types = FALSE)

```


# Introduction

You can and should cross-reference sections and sub-sections. 

The remainder of this paper is structured as follows. @sec-data, @sec-results, @sec-discussion, and @sec-conclusion.



# Data {#sec-data}

## Number of Crimes Committed in 2020 in Toronto Neighbourhoods

## Neighbourhood Profiles, 2021 (158 Model)

## Combining Crimes, Income and Population Data

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false
#| eval: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| eval: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("outputs/data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it. 

# Results {#sec-results}

## Toronto Neighbourhoods Crimes Committed in 2020 Statistics

## Toronto Neighbourhood Profile Based on 2021 Census Data

## Relationship Between Crimes Committed and Neighbourhood Population

## Relationship Between Crimes Committed and Neighbourhood Income




```{r}
#| echo: false
#| eval: false
#| label: tbl-modelresults
#| include: false
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```




# Discussion {#sec-discussion}

# Conclusion {#sec-conclusion}


\newpage


# References


